Optimal Decomposition-Recombination Plans for 2D Well-Constrained Graphs

### 2014-12-12 doc/GMP2015/s01_introduction.tex

\begin{itemize}

    \item (cross-section) of microtubules \cite{Necklace1} e.g., in ciliary membranes and transitions \cite{Necklace2}

    \item (cross-section) of hierarchical bone structure \cite{XX}

    \item interlinked cellulose or collagen microfibril monolayers e.g., in cell-walls \cite{CellWalls1} \cite{CellWalls1}

    \item more recent, engineered examples including  disordered graphene layers \cite{Graphene1} \cite{Graphene2} sometimes reinforced 
    by  microfibrils; and DNA assemblies \cite{Microfibrils1} including a recent Sierpinski carpet, bringing other structures 
    \cite{Microfibrils2} within reach.

    \item  silica bi-layers, glass \cite{SilicaGlass1} \cite{SilicaGlass2}, and materials that behave like assemblies of 
    2D particles under non-overlap constraints, e.g, jammed 
    disks on the plane \cite{JammedDisk1}

\end{itemize}

In order to study structural and mechanical properties such as rigidity, flexes and distribution of external stresses 
within the material layer, it is natural to model a material layer as a solution of geometric constraint systems 
arising from appropriate geometric primitives under metric or algebraic constraints. 
Such  2D {\bf{\em qusecs}} (quasi-uniform or self-similar constraint systems) can be used to 
to understand or design material layers (their solutions) with desired properties.

\subsection{Previous Work on Relevant 2D Geometric Constraint Systems}
We now briefly survey existing techniques for studying 2D {\em qusecs,}  many of which are
{\it bar-joint} systems (Examples 1,2 above), {\it multibody-pin} systems (Example 4,5) or {\it pinned-line incidence} systems (Example 3). 
The limitations of these techniques directly motivate the contributions of this paper.

\medskip\noindent\underbar{{\sl (1) Finding (vertex)-maximal generically rigid subsystems}}
Fast, graph-based algorithms exist (pebble-game \cite{XX}),   
for locating all maximal, generically rigid subsystems \cite{XX}. 
When the input is rigid, these algorithms do nothing, i.e, the output is the same as the input.
However, both for self-similar or just aperiodic 2D qusecs, it is imperative to recursively decompose rigid systems 
into their rigid subsystems, down to the level of geometric primitives, in order to 
understand or design properties at all scales, such as: rigidity, flexes, distribution 
of external stresses,  well as robustness/sensitivity to constraint variations.

\medskip\noindent\underbar{{\sl (2) Optimal Recursive Decomposition (DR-planning)}}
Recursive decomposition of geometric constraint systems has been formalized \cite{XX} and well-studied \cite{XX} 
as the {\sl Decomposition-Recombination (DR)-planning} problem.
For the above-mentioned classes of 2D qusecs, generic rigidity is 
a combinatorial property and hence each level of the decomposition should, in principle, be achievable by a graph-based algorithm as in (1), without
involving geometric information in the constraint system.  
Since many  such decompositions can exist for a given constraint system, 
criteria defining desirable or optimal DR-plans and DR-planning algorithms were given in \cite{XX}. An {\em optimal DR-plan} is one
that minimizes the maximum number of child subsystems of any parent system, a property that is 
crucial for obtaining a solution to the parent constraint system from the solutions of the child systems.
However, for general 2D qusecs, even when restricted to bar-joint systems, 
the optimal DR-planning problem was shown to be NP-hard \cite{XX}.

\medskip\noindent\underbar{{\sl (3) DR-plans for special classes and with other criteria}}
For a special class of 2D qusecs, namely  {\em tree-decomposable} systems \cite{XX} common in computer aided mechanical design, 
(which includes ruler-and-compass and Henneberg-I constuctible systems), all DR-plans turn out to 
be optimal. This satisfies the so-called {\em Church-Rosser} criterion, leading to highly efficient DR-planning algorithms.
Subsequently, alternate criteria were suggested   
such as {\em cluster minimality} 
requiring parent systems to be composed of  
a minimal set of at least 2 rigid proper subsystems (i.e., no proper subset forms a rigid system); 
and {\em proper maximality}, requiring child subsystems
to be maximal rigid proper subsystems of the parent system.
While polynomial time algorithms were given \cite{XX} to generate DR-plans meeting the cluster minimality criterion,
no such algorithm is known for the latter criterion.

\medskip\noindent\underbar{{\sl (4) Optimal Recombination and Solution Space Navigation}}
For fairly general 3D constraint systems, there are algorithms with formal guarantees that 
solve the parent (sub)systems that occur in the DR-plan \cite{XX, XX,XX}, 
provided the DR-plan meets some of the above criteria.
These algorithms  uncover underlying matroids to combinatorially optimize the algebraic complexity of (re)combining child system solutions, 
and algebraically solve the optimized recombination system to give a solution to the parent system at any
given level of the DR-plan. However, the generality of these algorithms trades-off against efficiency, and despite the optimization,
the best algorithms can still take
exponential time in the number of child subsystems in order to guarantee all solutions even for a single (sub)system in a DR-plan, 
and are prohibitively slow in practice. For the entire DR-plan, finding all solutions is barely tractable even with fast recombination algorithms
for the subsystems in the DR-plan. This is because even for the simplest systems, the problem is NP-hard \cite{XX} and there is a 
combinatorial explosion of solutions \cite{XX}.  Using the DR-plan and optimal recombination as a principled basis, high performance software exists \cite{XX,XX} to 
tame combinatorial explosion via user intervention.

\medskip\noindent\underbar{{\sl (5) Configuration Spaces of Underconstrained Systems}}
For underconstrained 2D bar-joint and multibody-pin cusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed \cite{XX, XX} that 
to complete them into isostatic systems \cite{XX, XX} and to find paths within the connected components \cite{XX,XX} of
standard cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{XX,XX}
to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation 
of connected components and paths.
These algorithms have decreasing efficiency as the subclass of systems gets bigger, starting with
underlying partial 2-tree graphs (alternately, tree-width 2, series-parallel, $K_4$ minor avoiding),  moving on   
to 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon and Cardioid), 
and further to general 1-dof tree-decomposable graphs. 
While software suites exist \cite{XX,XX}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.

\subsection{Contributions}
\label{sec:cont}
In this paper, we  navigate the NP-hardness barrier mentioned in (2) above, for finding optimal DR-plans by defining a 
so-called {\it canonical} DR-plan and showing a strong Church-Rosser property: {\it all canonical DR-plans for isostatic or underconstrained (independent) 
2D qusecs are optimal}. 
We then give an efficient ($O(n^2)$) algorithm to find a canonical (and hence optimal) DR-plan for all 3 types of 2D qusecs mentioned above.
canonical DR-plan 
illuminates the essence of the NP-hardness of finding optimal DR-plans for overconstrained  and gives classes of 
overconstrained 2DGMCS for which the above  algorithm finds an optimal DR-plan. We then show how to judiciously combine 
existing methods for recombination, search of so-called, convex Cayley configuration spaces, and navigation of realizations, 
to explore a connected component of the configuration space of a flexible 2DGMCS, given a starting configuration. 
Software implementation can be found at: \todo{Provide Link}.

The starting point of this paper is the observation that the above 2D Geometric Materials Constraint Systems (resp. 2D-GEMS) 
belong to a

However, these systems do not 
belong in the largest combinatorial class of 2D systems, whose configuration spaces have formally proven  characterizations and 
algorithmic guarantees (e.g. for algebraic complexity, path and volume measures, and  topology).

Specifically, these results \todo{[Citation Needed]} rely on the fact that rigid systems in this  class, in contrast to GMCS, 
have a straightforward and immediate {\sl optimal} recursive decomposition, (so-called {\sl optimal} DR-plan) and  recombination 
or realization (with a formal disambiguation by minimal orientation specification).

It is known \todo{[Citation Needed]} that already the optimal DR-planning problem for general 2DGMCS is NP-hard, intuitively 
because of a combinatorial explosion of polynomially-sized DR-plans.

\subsection{Previous work}

>>>>

\begin{enumerate}
  %  
    \item (cross-section) of microtubules \cite{Necklace1} e.g., in ciliary membranes and transitions \cite{Necklace2}
   % 
    \item (cross-section) of hierarchical bone structure \cite{XX}
    %
    \item interlinked cellulose or collagen microfibril monolayers e.g., in cell-walls \cite{CellWalls1} \cite{CellWalls1}
  %  
    \item more recent, engineered examples including  disordered graphene layers \cite{Graphene1} \cite{Graphene2} sometimes reinforced 
    by  microfibrils; and DNA assemblies \cite{Microfibrils1} including a recent Sierpinski carpet, bringing other structures 
    \cite{Microfibrils2} within reach.
    %
    \item  silica bi-layers, glass \cite{SilicaGlass1} \cite{SilicaGlass2}, and materials that behave like assemblies of 
    2D particles under non-overlap constraints, e.g, jammed 
    disks on the plane \cite{JammedDisk1}
%
\end{enumerate}
%
In order to study structural and mechanical properties such as rigidity, flexes and distribution of external stresses 
within the material layer, it is natural to model a material layer as a solution of geometric constraint systems 
arising from appropriate geometric primitives under metric or algebraic constraints. 
Such  2D {\bf{\em qusecs}} (quasi-uniform or self-similar constraint systems) can be used to 
to understand or design material layers (their solutions) with desired properties.
%
\subsection{Previous Work on Relevant 2D Geometric Constraint Systems}
We now briefly survey existing techniques for studying 2D {\em qusecs,}  many of which are
{\it bar-joint} systems (Examples 1,2 above), {\it multibody-pin} systems (Example 4,5) or {\it pinned-line incidence} systems (Example 3). 
The limitations of these techniques directly motivate the contributions of this paper.

\medskip\noindent\underbar{{\sl (i) Finding (vertex)-maximal generically rigid subsystems}}
Fast, graph-based algorithms exist (pebble-game \cite{XX}),   
for locating all maximal, generically rigid subsystems \cite{XX}. 
When the input is rigid, these algorithms do nothing, i.e, the output is the same as the input.
However, both for self-similar or just aperiodic 2D qusecs, it is imperative to recursively decompose rigid systems 
into their rigid subsystems, down to the level of geometric primitives, in order to 
understand or design properties at all scales, such as: rigidity, flexes, distribution 
of external stresses, boundary conditions for isostaticity, as well as behavior under constraint variations.

\medskip\noindent\underbar{{\sl (ii) Optimal Recursive Decomposition (DR-planning)}}
Recursive decomposition of geometric constraint systems has been formalized \cite{XX} and well-studied \cite{XX} 
as the {\sl Decomposition-Recombination (DR)-planning} problem.
For the above-mentioned classes of 2D qusecs, generic rigidity is 
a combinatorial property and hence each level of the decomposition should, in principle, be achievable by a graph-based algorithm as in (1), without
involving geometric information in the constraint system.  
Since many  such decompositions can exist for a given constraint system, 
criteria defining desirable or optimal DR-plans and DR-planning algorithms were given in \cite{XX}. An {\em optimal DR-plan} is one
that minimizes the maximum number of child subsystems of any parent system, a property that is 
crucial for obtaining a solution to the parent constraint system from the solutions of the child systems.
However, for general 2D qusecs, even when restricted to bar-joint systems, 
the optimal DR-planning problem was shown to be NP-hard \cite{XX}.

\medskip\noindent\underbar{{\sl (iii) DR-plans for special classes and with other criteria}}
For a special class of 2D qusecs, namely  {\em tree-decomposable} systems \cite{XX} common in computer aided mechanical design, 
(which includes ruler-and-compass and Henneberg-I constuctible systems), all DR-plans turn out to 
be optimal. This satisfies the so-called {\em Church-Rosser} criterion, leading to highly efficient DR-planning algorithms.
Subsequently, alternate criteria were suggested   
such as {\em cluster minimality} 
requiring parent systems to be composed of  
a minimal set of at least 2 rigid proper subsystems (i.e., no proper subset forms a rigid system); 
and {\em proper maximality}, requiring child subsystems
to be maximal rigid proper subsystems of the parent system.
While polynomial time algorithms were given \cite{XX} to generate DR-plans meeting the cluster minimality criterion,
no such algorithm is known for the latter criterion.

\medskip\noindent\underbar{{\sl (iv) Optimal Recombination and Solution Space Navigation}}
For fairly general 3D constraint systems, there are algorithms with formal guarantees that 
solve the parent (sub)systems that occur in the DR-plan \cite{XX, XX,XX}, 
provided the DR-plan meets some of the above criteria.
These algorithms  uncover underlying matroids to combinatorially optimize the algebraic complexity of (re)combining child system solutions, 
and algebraically solve the optimized recombination system to give a solution to the parent system at any
given level of the DR-plan. However, the generality of these algorithms trades-off against efficiency, and despite the optimization,
the best algorithms can still take
exponential time in the number of child subsystems in order to guarantee all solutions even for a single (sub)system in a DR-plan, 
and are prohibitively slow in practice. For the entire DR-plan, finding all solutions is barely tractable even with fast recombination algorithms
for the subsystems in the DR-plan. This is because even for the simplest systems, the problem is NP-hard \cite{XX} and there is a 
combinatorial explosion of solutions \cite{XX}.  Using the DR-plan and optimal recombination as a principled basis, high performance software exists \cite{XX,XX} to 
tame combinatorial explosion via user intervention.

\medskip\noindent\underbar{{\sl (v) Configuration Spaces of Underconstrained Systems}}
For underconstrained 2D bar-joint and multibody-pin cusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed \cite{XX, XX} that 
to complete them into isostatic systems \cite{XX, XX} and to find paths within the connected components \cite{XX,XX} of
standard cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{XX,XX}
to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation 
of connected components and paths.
These algorithms have decreasing efficiency as the subclass of systems gets bigger, starting with
underlying partial 2-tree graphs (alternately, tree-width 2, series-parallel, $K_4$ minor avoiding),  moving on   
to 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon and Cardioid), 
and further to general 1-dof tree-decomposable graphs. 
While software suites exist \cite{XX,XX}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.
%
\subsection{Contributions}
\label{sec:cont}
The contributions of this paper are the following.
\begin{itemize}
\item
We  navigate the NP-hardness barrier mentioned in (2) above, for finding optimal DR-plans by defining a 
so-called {\it canonical} DR-plan and showing a strong Church-Rosser property: {\it all canonical DR-plans for isostatic or underconstrained (independent) 
2D qusecs are optimal}. 
\item
We give an efficient ($O(n^2)$) algorithm to find a canonical (and hence optimal) DR-plan for all 3 types of 2D qusecs mentioned above.
The canonical DR-plan  
illuminates the essence of the NP-hardness of finding optimal DR-plans for overconstrained systems, and characterizes those 
overconstrained 2D qusecs for which the above algorithm does find an optimal DR-plan. 
\item
We show how to judiciously combine 
existing methods for recombination with searching  convex or low complexity Cayley configuration spaces to  navigate the solution space of isostatic 2D qusecs, and
explore connected components of the configuration space of underconstrained 2D qusecs. 
\item
We briefly describe applications of the above techniques to modeling and analyzing specific 2D material layers mentioned above.
\item
Software implementation and videos can be found at: \todo{Provide Link}.
\end{itemize}

%%%%%%%
XXXXXXXXXXXXXXXXXXXXXXx
